---
title: "Getting started with icpim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with icpim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(icpim)
```

The ICP Integrated Monitoring programme has measured whole forested catchments
since the late 1980s - deposition, soil, soil water, groundwater, runoff,
vegetation and biota at the same plots, year after year. The open dataset
([doi:10.5878/z376-2m63](https://doi.org/10.5878/z376-2m63)) publishes that
record as 21 CSV files, one per subprogramme, covering 55 sites in 14 countries
and about 1.2 million observations.

This package downloads those files, caches them, reads them with the right
column types, and decodes the code lists. It also handles a handful of
encoding traps that are easy to walk into and hard to notice afterwards.

## What is in the dataset

```{r}
im_subprogrammes
```

That table holds only what stays true between releases. Anything specific to
one version is asked for at the time, so it cannot go stale: `im_manifest()`
reads the current file list and sizes from the repository, and `im_coverage()`
measures rows, sites and year ranges from the data itself.

```{r, eval = FALSE}
im_manifest()        # file list and sizes for the pinned version
im_coverage("MC")    # what one subprogramme actually contains
```

Sites, with coordinates:

```{r}
head(im_sites)
```

## Reading data

`im_read()` is the entry point. The first call downloads and caches the file;
later calls read from the cache.

```{r, eval = FALSE}
pc <- im_read("PC", countries = "SE", years = 2000:2019)
```

Filters for `sites`, `countries`, `years` and `substances` are applied while
reading, which is worth using - the larger files run to 180,000 rows.

The examples below use the small extracts bundled with the package, so they
run offline:

```{r}
pc <- im_read_file(im_example("sample_PC.csv")) |> im_decode()
head(pc[, c("AREA", "SCODE", "date", "SUBST", "substance", "VALUE", "UNIT")])
```

Columns in UPPERCASE are exactly as published, so they match the ICP IM Manual
and the dataset README one for one. Columns in lowercase were added by this
package.

## Three traps worth knowing about

### Sodium's substance code is the string `NA`

This is the one that will cost you a day. In the ICP IM code list the
substance code for sodium is the two letters `NA`. Every default CSV reader in
R treats `"NA"` as a missing value, so:

```{r}
# What the obvious approach does to the code list:
tmp <- tempfile(fileext = ".csv")
write.csv(data.frame(SUBST = c("NA", "CL", "MG"), VALUE = 1:3), tmp,
          row.names = FALSE)
read.csv(tmp)$SUBST
```

Sodium has silently become a missing value, which then deletes every sodium
record and quietly breaks any `SUBST ==` comparison. The fix is
`na.strings = ""`, and `icpim` does this for you.

The same thing happened once before publication, and is a **known issue
affecting version 1 of the dataset only**: sodium's code arrives blank in
48,441 rows across twelve subprogrammes. It is corrected from version 2
onwards. `icpim` corrects it on read.

```{r}
sum(pc$SUBST == "NA")
unique(pc$substance[pc$SUBST == "NA"])
```

The `repair` argument controls this. Its default, `"auto"`, applies the
correction to version 1 and leaves later versions alone, so a script written
today keeps working unchanged when you move to version 2. `repair = FALSE`
gives you the file exactly as published; `repair = TRUE` forces the
correction.

A file that already carries the code needs no correction and is passed through
untouched whatever you set, so reading corrected and uncorrected releases
gives the same answer.

### Station codes are zero-padded

`SCODE` is a four-digit code such as `"0176"`. Read as a number it becomes
`176`, and it will no longer join or group correctly.

```{r}
unique(pc$SCODE)
```

`SCODE` distinguishes *collectors*, not replicates. A site can have more than
one gauge, running over different periods. Summing a flux over all of them
double-counts the overlap and manufactures a step where a collector starts or
stops, so check before aggregating:

```{r}
table(pc$SCODE, pc$year)[, 1:3]
```

### `FLAGSTA` is part of the key, not an annotation

In meteorology especially, the same site, level, month and parameter carries
several rows that differ only by their status flag: the monthly mean, the
minimum, the maximum, and the averaged daily extremes.

```{r}
am <- im_read_file(im_example("sample_AM.csv")) |> im_decode()
subset(am, SUBST == "TEMP" & LEVEL == 200 & YYYYMM == "200207",
       select = c(LEVEL, YYYYMM, FLAGSTA, stat, VALUE))
```

Five rows, one month, one sensor. Averaging without filtering mixes means with
extremes; across the whole dataset that biases mean air temperature by
+0.42 °C. For precipitation it is worse, because monthly sums (`S`) sit
alongside maximum *daily* sums (`SZ`).

Use the `stat` argument, which accepts either the codes or the readable names:

```{r, eval = FALSE}
temp <- im_read("AM", substances = "TEMP", stat = "mean")
```

`im_widen()` keeps `FLAGSTA` in the key by default, and refuses to pivot
rather than collapse duplicates silently:

```{r, error = TRUE}
im_widen(am, id_cols = c("AREA", "SCODE", "LEVEL", "YYYYMM"))
```

### Codes come from two lists, not one

`SUBST` and `PARAM` draw on two separate published code lists, and a companion
column says which: `LISTSUB` for `SUBST`, `PARLIST` for `PARAM`. `"DB"` means
the substance list, `"IM"` means the IM parameter list.

About 26,000 rows across seven subprogrammes use `IM` codes - `AOT40`,
`SOL_G`, `CEC_E`, `C/N`, `BDEN` - and none of them appear in the substance
list at all. Looking those up in `substance_codes.csv` alone silently returns
nothing.

The overlap matters too. 77 codes are in both lists, and one means different
things in each:

```{r}
im_codes("substance", "^ABS$")$name
subset(im_parameters, code == "ABS", select = c(subprog, name))$name
```

Both readings occur in `AL`, in the same file, told apart only by `PARLIST`.
`im_decode()` honours the discriminator, so you get the right name either way.

## Going wide

```{r}
w <- im_widen(pc)
head(w[, c("AREA", "date", "SO4S", "NO3N", "NA", "CL")])
```

Note that sodium's column is called `` `NA` ``, and needs backticks.

## Units and detection limits

Units are published per row and are not always consistent within a
subprogramme. Check before aggregating:

```{r}
head(im_units(pc), 8)
```

Values below the detection limit are published as the limit itself, flagged
`L`. Treating them as measurements biases means upward; dropping them biases
downward. `im_detection_limit()` makes the choice explicit rather than
implicit:

```{r}
halved <- im_detection_limit(pc, action = "half")
```

## Recording what you used

The version of `icpim` does not tell you which data an analysis used, and
cannot: `icpim.version` is a runtime option, so the same package version reads
any release. The record is attached to the data instead.

```{r}
pc <- im_read_file(im_example("sample_PC.csv"))
im_provenance(pc)
```

Read through `im_read()` it carries the release and its DOI as well. The record
survives `filter()`, `mutate()`, `select()`, `arrange()`, `im_widen()` and
`saveRDS()`, so it is still there after reshaping and after a saved object is
reloaded. It is dropped by `summarise()` and by joins, which build a genuinely
new table; capture it first if you need it afterwards.

## Reproducibility

The deposit is updated annually, and each release is a new version with its
own DOI. `icpim` pins version 1 so that a script keeps returning the same
numbers:

```{r}
im_version()
im_doi()
```

Each release has its own DOI, and these follow the pinned version, so what you
cite is always what you read. Record both in your output.

`im_check_version()` reports whether a newer release has appeared. The package
never follows it on its own: annual updates revise past values as well as
adding new ones, so moving would change the numbers an existing analysis
returns. Move deliberately with `options(icpim.version = "2")`. The cache is
keyed by version, so files from two releases are never mixed.

The data are CC BY 4.0, which requires attribution:

```{r}
im_cite()
```

## Where files are stored

```{r, eval = FALSE}
im_cache_dir()   # under tools::R_user_dir("icpim", "cache")
im_cache_list()  # what is cached
im_cache_clear() # remove it
```

Set `options(icpim.cache_dir = "data-raw/icpim")` to keep an analysis
self-contained, or `im_download("all")` to fetch everything (about 95 MB)
before going offline.
