---
title: "Getting started with icpim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with icpim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(icpim)
```

The ICP Integrated Monitoring programme has measured whole forested catchments
since the late 1980s — deposition, soil, soil water, groundwater, runoff,
vegetation and biota at the same plots, year after year. The open dataset
([doi:10.5878/z376-2m63](https://doi.org/10.5878/z376-2m63)) publishes that
record as 21 CSV files, one per subprogramme, covering 55 sites in 14 countries
and about 1.2 million observations.

This package downloads those files, caches them, reads them with the right
column types, and decodes the code lists. It also handles a handful of
encoding traps that are easy to walk into and hard to notice afterwards.

## What is in the dataset

```{r}
im_subprogrammes[, c("subprog", "name", "n_rows", "n_sites")]
```

Sites, with coordinates:

```{r}
head(im_sites)
```

## Reading data

`im_read()` is the entry point. The first call downloads and caches the file;
later calls read from the cache.

```{r, eval = FALSE}
pc <- im_read("PC", countries = "SE", years = 2000:2019)
```

Filters for `sites`, `countries`, `years` and `substances` are applied while
reading, which is worth using — the larger files run to 180,000 rows.

The examples below use the small extracts bundled with the package, so they
run offline:

```{r}
pc <- im_read_file(im_example("sample_PC.csv")) |> im_decode()
head(pc[, c("AREA", "SCODE", "date", "SUBST", "substance", "VALUE", "UNIT")])
```

Columns in UPPERCASE are exactly as published, so they match the ICP IM Manual
and the dataset README one for one. Columns in lowercase were added by this
package.

## Three traps worth knowing about

### Sodium's substance code is the string `NA`

This is the one that will cost you a day. In the ICP IM code list the
substance code for sodium is the two letters `NA`. Every default CSV reader in
R treats `"NA"` as a missing value, so:

```{r}
# What the obvious approach does to the code list:
tmp <- tempfile(fileext = ".csv")
write.csv(data.frame(SUBST = c("NA", "CL", "MG"), VALUE = 1:3), tmp,
          row.names = FALSE)
read.csv(tmp)$SUBST
```

Sodium has silently become a missing value, which then deletes every sodium
record and quietly breaks any `SUBST ==` comparison. The fix is
`na.strings = ""`, and `icpim` does this for you.

In the published files this has already happened once, upstream of
publication: sodium's code arrives blank in 48,441 rows across twelve
subprogrammes. `im_read()` restores it by default.

```{r}
sum(pc$SUBST == "NA")
unique(pc$substance[pc$SUBST == "NA"])
```

Pass `repair = FALSE` to get the file exactly as published.

The evidence that these rows really are sodium: their count tracks chloride
and magnesium to within about 1% in every water-chemistry file; their units
are right for sodium in each medium (mg/l in waters, mg/kg in tissue, meq/kg
exchangeable in soil); and pairing them with chloride at the same site and
month gives a mass ratio of 1.83 in throughfall against seawater's 1.80.
Runoff gives 0.86, which is the expected signature of silicate weathering
adding sodium without chloride.

### Station codes are zero-padded

`SCODE` is a four-digit code such as `"0176"`. Read as a number it becomes
`176`, and it will no longer join or group correctly.

```{r}
unique(pc$SCODE)
```

`SCODE` distinguishes *collectors*, not replicates. A site can have more than
one gauge, running over different periods. Summing a flux over all of them
double-counts the overlap and manufactures a step where a collector starts or
stops, so check before aggregating:

```{r}
table(pc$SCODE, pc$year)[, 1:3]
```

### `FLAGSTA` is part of the key, not an annotation

In meteorology especially, the same site, level, month and parameter carries
several rows that differ only by their status flag: the monthly mean, the
minimum, the maximum, and the averaged daily extremes.

```{r}
am <- im_read_file(im_example("sample_AM.csv")) |> im_decode()
subset(am, SUBST == "TEMP" & LEVEL == 200 & YYYYMM == "200207",
       select = c(LEVEL, YYYYMM, FLAGSTA, stat, VALUE))
```

Five rows, one month, one sensor. Averaging without filtering mixes means with
extremes; across the whole dataset that biases mean air temperature by
+0.42 °C. For precipitation it is worse, because monthly sums (`S`) sit
alongside maximum *daily* sums (`SZ`).

Use the `stat` argument, which accepts either the codes or the readable names:

```{r, eval = FALSE}
temp <- im_read("AM", substances = "TEMP", stat = "mean")
```

`im_widen()` keeps `FLAGSTA` in the key by default, and refuses to pivot
rather than collapse duplicates silently:

```{r, error = TRUE}
im_widen(am, id_cols = c("AREA", "SCODE", "LEVEL", "YYYYMM"))
```

## Going wide

```{r}
w <- im_widen(pc)
head(w[, c("AREA", "date", "SO4S", "NO3N", "NA", "CL")])
```

Note that sodium's column is called `` `NA` ``, and needs backticks.

## Units and detection limits

Units are published per row and are not always consistent within a
subprogramme. Check before aggregating:

```{r}
head(im_units(pc), 8)
```

Values below the detection limit are published as the limit itself, flagged
`L`. Treating them as measurements biases means upward; dropping them biases
downward. `im_detection_limit()` makes the choice explicit rather than
implicit:

```{r}
halved <- im_detection_limit(pc, action = "half")
```

## Reproducibility

The deposit is updated annually, and each release is a new version with its
own DOI. `icpim` pins version 1 so that a script keeps returning the same
numbers:

```{r}
im_version()
im_doi()
```

Move deliberately with `options(icpim.version = "2")`. The cache is keyed by
version, so files from two releases are never mixed.

The data are CC BY 4.0, which requires attribution:

```{r}
im_cite()
```

## Where files are stored

```{r, eval = FALSE}
im_cache_dir()   # under tools::R_user_dir("icpim", "cache")
im_cache_list()  # what is cached
im_cache_clear() # remove it
```

Set `options(icpim.cache_dir = "data-raw/icpim")` to keep an analysis
self-contained, or `im_download("all")` to fetch everything (about 95 MB)
before going offline.
